Mechanising Recursion Schemes with Magic-Free Coq Extraction

### 2024-10-10 paper/paper.tex
related work

Abreu et al.~\cite{AbreuDHJMS23} encode divide-and-conquer computations in Coq,
using a recursion scheme in which termination is entirely enforced by its
typing.  This is a significant advance, since it avoids \emph{completely} the
need for termination proofs. Their work differs from ours in that they require
the functional extensionality axiom, and the use of impredicative \coq{Set}.
The authors justify well the use of impredicative \coq{Set} and its
compatibility with the functional extensionality axiom. In contrast, our
development remains entirely \emph{axiom-free}. Another key difference with our
approach is that they do not discuss what the resulting extracted code
\emph{looks like} (that is, whether the extracted OCaml code resembles the
natural formulation of the recursive function). Through experiments, we found
that their formalisation leads to \ocaml{Obj.magic}, and code with a complex
structure that may be hard to understand or interface with other OCaml code.
Due to the great benefit of entirely avoiding termination proofs, it would be
interesting to extend their approach to improve code extraction.

\subsection{Termination Checking}\label{sec:termcheck}

Termination and productivity are non-trivial properties of programs, and various
methods have been proposed for checking that these properties hold. A common
approach is guardedness checking~\cite{Eduardo94}, which is a
\emph{syntactic} check that definitions avoid the introduction of
non-normalizable terms. This sort of check generally looks for a
\emph{structural decrease} of arguments in the recursive calls of a function.
Coq uses such a check, and it works for many classic functional programming
patterns (like \haskell{map} and \haskell{foldr}). However, some desirable
definitions will not be accepted by such checks.

In particular, the problem of \emph{nonstructural recursion} (including
divide-and-conquer algorithms) is well-studied~\cite{BoveKS16}.
% A distinction between standard and
% nonstandard recursion lies in the fact that standard recursion adheres strictly
% to recursing on pattern variables, avoiding reliance on results from other
% function calls. Nonstandard recursion may deviate from this convention, allowing
% for more flexibility.
Certain functions that are not structurally recursive can be reformulated using
a nonstandard approach to achieve structural recursion~\cite{AbreuDHJMS23}.
Take, for instance, division by iterated subtraction, which is inherently
non-structurally recursive since it involves recursion based on the result of a
subtraction. There is a nonstandard implementation of divivion found on Coq's
standard library, which involves a four-argument function that effectively
combines subtraction and division. Similarly, the mergesort in Coq's standard
library uses an ``explicit stack of pending merges'' in order to avoid issues
with nonstructural definitions. There is a major downside, however; as noted by
Abreu et al., the result is ``barely recognizable as a form of
mergesort''~\cite{AbreuDHJMS23}.
There are common approaches in Coq to deal with termination, but none of these
approaches address program calculation techniques, and the mechanisation of
fusion laws.

@@

\emph{Deforestation}, introduced by Wadler~\cite{wadler88}, is a specific
approach to fusion that aims to remove intermediate trees (such as lists)
created between chained operations. A well-known method of deforestation is
\haskell{foldr/build} fusion~\cite{gill93}, which optimizes list-processing by
fusing list-producing functions (like \haskell{map} or \haskell{filter}) with
list-consuming ones (like \haskell{foldr}), avoiding the creation of temporary
lists.
%
Extending this idea, stream fusion generalizes the concept to handle other data
types, using a lazy stream abstraction to represent sequences and fuse
transformations like \haskell{map} and \haskell{fold}~\cite{coutts07}. Both
approaches eliminate redundant structures and ensure computations happen in a
single efficient pass over the data.

Another dimension of fusion is \emph{tupling}, a technique used to combine
multiple recursive functions into a single function by grouping their results
into a tuple~\cite{hughes89}. This approach helps avoid redundant traversals of
the same data structure when multiple operations are applied.
% In addition to
% techniques like \haskell{foldr/build} fusion and stream fusion, tupling
% complements deforestation by allowing multiple recursive functions to share a
% single traversal, further optimizing computations and minimizing intermediate
% data structures.

>>>>

Abreu et al.~\cite{AbreuDHJMS23} encode divide-and-conquer computations in Coq,
using a recursion scheme in which termination is entirely enforced by its
typing.  This is a significant advance, since it avoids \emph{completely} the
need for termination proofs. Their work differs from ours in that they require
the functional extensionality axiom, and the use of impredicative \coq{Set}.
The authors justify well the use of impredicative \coq{Set} and its
compatibility with the functional extensionality axiom. In contrast, our
development remains entirely \emph{axiom-free}. Another key difference with our
approach is that they do not discuss what the resulting extracted code
\emph{looks like}.
% (that is, whether the extracted OCaml code resembles the
% natural formulation of the recursive function).
Through experiments, we found that their formalisation leads to
\ocaml{Obj.magic}, and code with a complex structure that may be hard to
understand or interface with other OCaml code. Due to the great benefit of
entirely avoiding termination proofs, it would be interesting to extend their
approach to improve code extraction.

\subsection{Termination Checking of Nonstructural Recursion}\label{sec:termcheck}

% Termination and productivity are non-trivial properties of programs, and various
% methods have been proposed for checking that these properties hold. A common
% approach is guardedness checking~\cite{Eduardo94}, which is a
% \emph{syntactic} check that definitions avoid the introduction of
% non-normalizable terms. This sort of check generally looks for a
% \emph{structural decrease} of arguments in the recursive calls of a function.
% Coq uses such a check, and it works for many classic functional programming
% patterns (like \haskell{map} and \haskell{foldr}). However, some desirable
% definitions will not be accepted by such checks.

%In particular,
The problem of \emph{nonstructural recursion} (including
divide-and-conquer algorithms) is well-studied~\cite{BoveKS16}.
% A distinction between standard and
% nonstandard recursion lies in the fact that standard recursion adheres strictly
% to recursing on pattern variables, avoiding reliance on results from other
% function calls. Nonstandard recursion may deviate from this convention, allowing
% for more flexibility.
Certain functions that are not structurally recursive can be reformulated using
a nonstandard approach to achieve structural recursion~\cite{AbreuDHJMS23}.
Take, for instance, division by iterated subtraction, which is inherently
non-structurally recursive since it involves recursion based on the result of a
subtraction. There is a nonstandard implementation of divivion found on Coq's
standard library, which involves a four-argument function that effectively
combines subtraction and division. Similarly, the mergesort in Coq's standard
library uses an ``explicit stack of pending merges'' in order to avoid issues
with nonstructural definitions. There is a major downside, however; as noted by
Abreu et al., the result is ``barely recognizable as a form of
mergesort''~\cite{AbreuDHJMS23}.
There are common approaches in Coq to deal with termination, but none of these
approaches address program calculation techniques, and the mechanisation of
fusion laws.

@@

\emph{Deforestation}, introduced by Wadler~\cite{wadler88}, is a specific
approach to fusion that aims to remove intermediate trees (such as lists)
created between chained operations. A well-known method of deforestation is
\haskell{foldr/build} fusion~\cite{gill93}, which optimizes list-processing by
fusing list-producing functions (like \haskell{map} or \haskell{filter}) with
list-consuming ones (like \haskell{foldr}), avoiding the creation of temporary
lists.
%
Extending this idea, stream fusion generalizes the concept to handle other data
types, using a lazy stream abstraction to represent sequences and fuse
transformations like \haskell{map} and \haskell{fold}~\cite{coutts07}.
% Both approaches eliminate redundant structures and ensure computations happen in
% a single efficient pass over the data.

% Another dimension of fusion is \emph{tupling}, a technique used to combine
% multiple recursive functions into a single function by grouping their results
% into a tuple~\cite{hughes89}. This approach helps avoid redundant traversals of
% the same data structure when multiple operations are applied.
% In addition to
% techniques like \haskell{foldr/build} fusion and stream fusion, tupling
% complements deforestation by allowing multiple recursive functions to share a
% single traversal, further optimizing computations and minimizing intermediate
% data structures.

In the context of Coq, Danvy explores the use of fold/unfold lemmas to mechanize
equational reasoning about recursive programs in the Coq proof
assistant~\cite{DANVY_2022}. By formalizing recursive program transformations,
fold-unfold lemmas enable optimizations like fusion by allowing programs to be
systematically simplified, eliminating intermediate structures.
